Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-28 MScDiplom_Lemenkova.tex
updated Valentijn's comments: 1,2,4,6,7,9,11,12 (8 from total 15), revised research objectives and aims, Latin words, structure of paragraphs

\begin{center}
\textbf{Seagrass mapping and monitoring along the coasts of Crete, Greece}\\
 {by Lemenkova Polina}
\end{center}

@@

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair)
	\item Name Examiner 2
	\item External Examiner
\end{itemize}

@@

\textbf{Member Supervisors}: 
\begin{itemize}
	\item[*] Venus V. (Department of Natural Resource, ITC)
	\item[*] Toxopeus A.G. (Department of Natural Resource, ITC)
\end{itemize}

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using Gdal technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to marine environment, they are however, more closely related to the lily family
(Liliaceae) than to the true grass, despite their name “seagrass”, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}.\\
Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

and is widely spread along the coasts of Crete \cite{Dumay02} \label{Dumay02}. 
It plays an important role in a number
of geomorphological and ecological processes. Namely, it is a source of food for herbivorous fauna as
well as shelter zones for fish and other marine organisms; it contributes to the nutrient recycling; it
provides sediments stability by reducing the degree of water movements, etc \cite{Francour99}\label{Francour99}.\\
The purpose of current MSc research work aims to apply methods of remote sensing and GIS
including processing and classification of satellite, aerial photos and videometric underwater imaging
as well as GIS-based spatial analysis towards mapping and environmental monitoring of \emph{P. oceanica}
along the coasts of Crete island, Greece. The technical implementation is based on ENVI, Erdas
Imagine, WASI and ArcGIS software using aerial and satellite images as well as results of the
underwater videometric measurements.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate are--as), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

\subsubsection{Ecological significance of the seagrasses}
Seagrass plays vital role in the marine ecosystems of the world ocean. Seagrasses are the only
flowering plant in the world that is able to live completely submerged. Seagrass is a habitat for
numerous marine fish species \cite{Nagelkerken00}\label{Nagelkerken00}, source of primary production and food for
fish, turtles and other organisms, which gives them special environmental value \cite{Noralez10}\label{Noralez10}.\\
Seagrass meadows produce enormous quantities of organic matter (leaves, epiphytes), which
constitutes the basis of the food web both within and outside the ecosystem \cite{Gobert06}\label{Gobert06}.

\begin{wrapfigure}{r}{0.4\textwidth}
\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}

Finally, seagrasses are an important component in the environmental "food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.\\
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine “algae”, “seagrasses” are vastly different from them. There are
fundamental differences between both marine organisms, the major of them should be briefly
mentioned: first, seagrasses are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas algae are simple organisms that can only holdfast; secondly, seagrasses are
complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while algae
have simple few cell structure with spores and gametes; finally, seagrasses uptake nutrient through
root system, while algae nourish directly from the water column \cite{Dixon05}\label{Dixon05}. There are other
differences between “seagrass” and “algae” but they go beyond the scope of this work.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
This naturally increase environmental pressing and leads to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 percent of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.

\subsubsection{General characteristics of \textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-6.jpg}
	\caption{\textit{	extit{P.oceanica}}.\\
	Source \cite{Luque04}\label{Luque04}}
	\label{fig:6}
\end{wrapfigure}

@@

The in-situ fieldwork including underwater videographic measurements is an important part of the

@@

\subsection{Research objective}
The current MSc research aims to explore the environmental conditions and dynamics of the spatial
distribution of \textit{P.oceanica} seagrass meadows along the Cretan coast, based on the remote sensing and
GIS techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources:
i) spectra of 	extit{P.oceanica}, carbonate sand, silt and other seafloor types
ii) satellite imagery: Landsat, ASTER, MODIS, MERIS
iii) aerial photos: Google Earth
iv) in-situ fieldwork data of underwater videographic measurements
v) vector GIS layers.
The research aims to explore the dynamics of the environmental changes of the \textit{P.oceanica} meadows
over time (2000-2010), as well as the extent of the spatial distribution over the whole island. The
main research objective is monitoring the seagrass \textit{P.oceanica} along the coasts of Crete, in order to
analyse the environmental changes of the 	extit{P.oceanica} meadows over the last 10 years (2000-2010)
and the spatial extent of its current distribution.
This research is supported by the in-situ measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: underwater videometric measurements made by Olympus camera, aerial
Google Earth and satellite images from different sources, spatial GIS and statistical analysis.

\subsubsection{General objective}
The main objective of this study is to investigate the distribution of seagrass along the northern coast
of Crete using remote sensing techniques.
General objectives:
1) Mapping the spatial distribution of the seagrass \textit{P.oceanica} along the northern coasts of
Crete Island
2) Monitoring the dynamics of the environmental changes of the \textit{P.oceanica} meadows over
time (2000-2010) in Ligaria beach, Agia Pelagia district.

\subsubsection{Specific objectives}
1. To apply remote sensing data (Google Earth aerial images, Landsat, SPOT satellite images)
for the monitoring of the seagrass meadows distribution
2. To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor
cover types (sand and silt) using WASI water colour simulation software
3. To use methods of the in situ diving observations and underwater videometric measurements
by Olympus camera in order to receive large-scale imagery of the 	extit{P.oceanica} mattes
4. To perform images classification (supervised and unsupervised) for the thematic mapping of
the 	extit{P.oceanica} seagrass distribution along the coasts of northern Crete.

\subsection{Research questions}
1. Is \textit{P.oceanica} spectrally distinct from other seafloor cover types with varying in-situ
conditions ?
2. Whether broadband and hyperspectral remote sensing data can be used for the mapping of
\textit{P.oceanica} ?

\subsection{Hypotheses testing}
A statistical testing will be used to compare between the spectral responses of the different seagrass
types, whether it is spectrally distinct and at least one pair is statistically different at every spectral
band.
Hypothesis Ho: seagrass types are not spectrally distinct from other seafloor types with
varying in-situ conditions, which means
Ho: μ1 =μ2 =μ3=...= μn.
The alternative Hypothesis Ha claims the opposite statement: seagrass is spectrally
distinct with varying in-situ conditions, Ho: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn.
The distribution of the spectral responses at every spectral band is assumed to be normal as well as
the equality of the statistical variances.
The hypothesis testing is suggested to be carried out using the ANOVA statistical test. The purpose of
ANOVA test is to visualize in an effective and quick way the spectral differences between seagrass
species and their spatial distribution. Thus, the key hypotheses of the research will be tested to prove
whether the results of the research are accurate, reasonable and correct.

>>>>

\begin{center}
\textbf{Seagrass mapping and monitoring along the coasts of Crete, Greece}\\
\vspace{1em}
 {by Lemenkova Polina}
\end{center}

@@

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair)
	\item Name Examiner 2
	\item External Examiner \hfill{Dr. \textit{Petter Pilesj\"o} (Lund University, GIS Centre)}
\end{itemize}

@@

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] Mr. \textit{Valentijn Venus} (Department of Natural Resource, ITC)
	\item[* ] Dr. \textit{Albertus G. Toxopeus} (Department of Natural Resource, ITC)
\end{itemize}

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using Gdal technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

and is widely spread along the coasts of Crete \cite{Dumay02} \label{Dumay02}. 
It plays an important role in a number
of geomorphological and ecological processes. Namely, it is a source of food for herbivorous fauna as
well as shelter zones for fish and other marine organisms; it contributes to the nutrient recycling; it
provides sediments stability by reducing the degree of water movements, etc \cite{Francour99}\label{Francour99}.\\
The purpose of current MSc research work aims to apply methods of remote sensing analysis, including Radiative Transfer Models (RTM), 
GIS-based spatial analysis, processing and classification of satellite and aerial photos, as well as videometric underwater footage, towards mapping and environmental monitoring of seagrass \textit{P. oceanica} along the selected locations of the northern coasts of Crete island, Greece. The technical implementation is based on WASI RTM software, GIS (ILWIS, Erdas Imagine and ArcGIS), using aerial and satellite images and the results of the underwater videometric measurements.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

\subsubsection{Ecological significance of the seagrasses}
Seagrass plays vital role in the marine ecosystems of the world ocean. 
\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}
Seagrasses are the only flowering plant in the world that is able to live completely submerged. Seagrass is a habitat for
numerous marine fish species \cite{Nagelkerken00}\label{Nagelkerken00}, source of primary production and food for
fish, turtles and other organisms, which gives them special environmental value \cite{Noralez10}\label{Noralez10}.\\
Seagrass meadows produce enormous quantities of organic matter (leaves, epiphytes), which
constitutes the basis of the food web both within and outside the ecosystem \cite{Gobert06}\label{Gobert06}.
Finally, seagrasses are an important component in the environmental "food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.\\
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine “algae”, “seagrasses” are vastly different from them. There are
fundamental differences between both marine organisms, the major of them should be briefly
mentioned: first, seagrasses are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas algae are simple organisms that can only holdfast; secondly, seagrasses are
complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while algae
have simple few cell structure with spores and gametes; finally, seagrasses uptake nutrient through
root system, while algae nourish directly from the water column \cite{Dixon05}\label{Dixon05}. There are other
differences between “seagrass” and “algae” but they go beyond the scope of this work.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.

@@

Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 \% of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.

\subsubsection{General characteristics of \textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.
The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-1-6.jpg}
	\caption{\textit{P.oceanica}.\\
	Source \cite{Luque04}\label{Luque04}}
	\label{fig:6}
\end{wrapfigure}

@@

The \textit{in-situ} fieldwork including underwater videographic measurements is an important part of the

@@

\subsection{Research objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and GIS techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: Landsat TM, Landsat ETM+
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
The research aims to explore the limitations of the application of the Radiative Transfer Models (further RTMs) and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the RTM Water Colour Simulator WASI,  underwater videometric measurements made by Olympus camera, aerial Google Earth and satellite images from different sources, spatial GIS and statistical analysis.

\subsubsection{General objective}
The main objective of this study is to analyse the optical properties of the seagrass \textit{P.oceanica} and other seafloor types, and to apply the remote sensing techniques towards the investigation of the seagrass distribution in the selected locations along the northern coast of Crete.
General objectives:
	\begin{itemize}
		\item[*]Analysing spectral reflectance of the \textit{P.oceanica} and other seafloor cover types means of radiative transfer model tools (RTMs), using WASI.
		\item[*]Mapping spatial distribution of the seagrass \textit{P.oceanica} over selected locations along the northern coasts of Crete Island 
	\end{itemize}

\subsubsection{Specific objectives}
	\begin{itemize}	
		\item[]1.	To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor cover types (sand and silt) using WASI water colour simulation software 
		\item[]2.	To use methods of the in situ diving observations and underwater videometric measurements by Olympus camera in order to receive large-scale imagery of the P.oceanica mattes 
		\item[]3.	To apply remote sensing data (Google Earth aerial images, Landsat TM and ETM+ satellite images) for the monitoring of the seagrass meadows distribution 
		\item[]4.	To perform supervised images classification for the thematic mapping of the \textit{P.oceanica} seagrass distribution along the selected locations over the coasts of northern Crete. 
	\end{itemize}

\subsection{Research questions}
	\begin{itemize}	
		\item Is \textit{P.oceanica} spectrally distinct from other seafloor cover types with varying \textit{in-situ} conditions ?
		\item Whether broadband and hyperspectral remote sensing data can be used for the mapping of \textit{P.oceanica} ?
	\end{itemize}

\subsection{Hypotheses testing}
A statistical testing will be used to compare between the spectral responses of the different seafloor cover types (i.e. sand and \textit{P. oceanica}),
 whether it is spectrally distinct and at least one pair is statistically different at every spectral
band.
Hypothesis Ho: seagrass types are not spectrally distinct from other seafloor types with
varying \textit{in-situ} conditions, which means
Ho: μ1 =μ2 =μ3=...= μn.
The alternative Hypothesis Ha claims the opposite statement: seagrass is spectrally
distinct with varying \textit{in-situ} conditions, Ho: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn.
The distribution of the spectral responses at every spectral band is assumed to be normal as well as
the equality of the statistical variances.
The hypothesis testing is suggested to be carried out using the ANOVA statistical test. The purpose of
ANOVA test is to visualize in an effective and quick way the spectral differences between seagrass
species and their spatial distribution. Thus, the key hypotheses of the research will be tested to prove
whether the results of the research are accurate, reasonable and correct.
